Neural Power Unit \\-- A Bayesian Approach To Neural Arithmetic

### 2020-05-20 papers/main.tex
paper/taylor updates

\newcommand{\real}{\text{real}}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{compare-npu-nmu-nalu-xy-xdivy.png}
  \caption{Comparison of different models learning the function
  $(x,y)\rightarrow(xy,x/y)$ with $(x,y)\in\mathcal U^2(0.1,2)$.  Each
  plot shows the mean-squared error on a logarithmic color scale ($\hat{\bm
  t}=\text{model}(\bm x)$).  The Dense network learns both multiplication and
  division, but fails to extrapolate. NALU learns both tasks perfectly, but
  completely fails if one of the inputs is negative.  The NMU successfully
  learns multiplication, but not division. Only the NPU learns both tasks and
  can extrapolate with an error that is less than $0.01$ in most places.}%
  \label{fig:compare_npu_nmu_nalu_xy_xdivy}
\end{figure}
\todo{add in depth experiment desc. of fig.~\ref{fig:compare_npu_nmu_nalu_xy_xdivy} in appendix}
\todo{add sqrt to table~\ref{tab:compare_npu_npu_nalu}?}
\begin{table}
  \centering
  \caption{Comparison of different $2\times2$ layers on the task
  $(x,y)\rightarrow(xy,\frac{x}{y})$.  $(x,y) \in \mathcal U^2(-2,2)$.}
  \label{tab:compare_npu_npu_nalu}
  \input{table-x*y-xdivy-pos.tex}
\end{table}

\section{Neural Power Unit}%
\label{sec:neural_power_unit}

Eq.~\ref{eq:nalu_mult} enables the NALU to learn multiplication. If the weight
matrix $\bm W$ would not be restricted to values in the range $(-1,1)$ it could
in principle learn arbitrary power functions, because
\begin{align}
  z = \exp(w\log x) = x^w \text{ for } x>0.
\end{align}
However, this would complicate the convergence issues of the NALU even more.
We therefore suggest to leverage the power of the complex plane, by using the
complex logarithm and by promoting $w$ to a complex number.  This will enable
us not only to lift the positivity constraint on $x$, but also provides and
additional dimension that helps to avoid regions with an uninformative gradient
signal\todo{rephrase this. do we have an educating example where it is obvious
that complex plane helps? the stuff in fig.1 does not work as well with
non-complex NPU, but I have not idea how to plot that}.
Introducing a complex weight matrix somewhere in a larger network would mean that
the gradients in other layers would potentially become complex as well, which
would result in converting the whole network to complex numbers.
This seems like a high price to pay, but fortunately there is a way around this.
We can output only the real part of the result $z$, but still maintain a complex $w$.

\newcommand{\wre}{w_{\text{re}}}
\newcommand{\wim}{w_{\text{im}}}
\begin{align}
  \real(z) &= \real(\exp(w\log x)) \\
    &= \real(\exp((\wre + i\wim) (\log r + ik\pi))) \\
    &= \exp(\wre\log r - \wim k\pi) \real(\exp(i\wim\log r + ik\pi\wre)) \\
    &= \exp(\wre\log r - \wim k\pi) \cos(\wim\log r + k\pi\wre)
\end{align}
\todo{explicitly compute gradients?}

Our \emph{Neural Power Unit} (NPU) can learn arbitrary power functions (which
includes division) while still being able to correctly deal with negative
inputs.  The NPU layer is defined by
\begin{align}
  k_i &= \begin{cases}
     0  & x_i \leq 0 \\
    \pi & x_i > 0
  \end{cases} \\
  \bm r &= |\bm x| \\
  \bm m &= \exp(\bm M \log(\bm r)) \odot \cos(\bm M \bm k)
\end{align}
where $\bm k$ is a vector that is zero where $\bm x$ is positive and $\pi$
where it is negative, $\bm r$ is the elementwise absolute value, and $\bm M$
the multiplication matrix that we want to learn.

The NPU is inspired by the multiplication part of the NALU
Eq.~\ref{eq:nalu_mult}, extended to negative inputs by taking advantage of the
complex logarithm. In general, for any complex number $z$\footnote{Note that
complex numbers can be represented in the polar, complex plane where
$z=re^{i\varphi}$ with $r>0$ and $\varphi \in (0,2\pi)$.}
\begin{align}
  \log(z) = \log\left(r\cdot e^{i\varphi}\right)
     = \log(r) + i\varphi.
\end{align}
However, as $z$ is assumed to be real, we can simplify as follows:
\begin{align}
  \log(z) = \log(r) + ik\pi,
\end{align}
where $k \in [0,1]$ is zero for positive inputs and one for negative inputs $z$.

\begin{align}
  \bm m &= \exp(\bm M \log(\bm x)) \\
    &= \exp(\bm M ( \log(\bm r) + i\pi\bm k ))
\end{align}

\begin{align}
  \bm m_{\text{re}} &= \real(\exp(\bm M \log \bm r + i\pi\bm M\bm k)) \\
    &= \real(\exp(\bm M \log \bm r) \odot (\cos(\pi \bm M \bm k) + i \sin(\pi \bm M \bm k))) \\
    &= \exp(\bm M \log \bm r) \odot \cos(\pi \bm M \bm k)
\end{align}

@@

>>>>

\newcommand{\real}{\text{Re}}
\newcommand{\imag}{\text{Im}}
\newcommand{\wre}{w_{r}}
\newcommand{\wim}{w_{i}}
\newcommand{\Wre}{\bm W_{r}}
\newcommand{\Wim}{\bm W_{i}}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{compare-npu-nmu-nalu-xy-xdivy.png}
  \caption{Comparison of different models learning the function
  $(x,y)\rightarrow(xy,x/y)$ with $(x,y)\in\mathcal U^2(0.1,2)$.  Each
  plot shows the mean-squared error on a logarithmic color scale ($\hat{\bm
  t}=\text{model}(\bm x)$).  The Dense network learns both multiplication and
  division, but fails to extrapolate. NALU learns both tasks perfectly, but
  completely fails if one of the inputs is negative.  The NMU successfully
  learns multiplication, but not division. Only the NPU learns both tasks and
  can extrapolate with an error that is less than $0.01$ in most places.}%
  \label{fig:compare_npu_nmu_nalu_xy_xdivy}
\end{figure}
\todo{add in depth experiment desc. of fig.~\ref{fig:compare_npu_nmu_nalu_xy_xdivy} in appendix}

\section{Neural Power Units}%
\label{sec:neural_power_unit}

Eq.~\ref{eq:nalu_mult} enables the NALU to learn multiplication. If the weight
matrix $\bm W$ would not be restricted to values in the range $(-1,1)$ it could
in principle learn arbitrary power functions, because
\begin{align}
  z = \exp(w\log x) = x^w \text{ for } x>0.
\end{align}

However, this would complicate the convergence issues of the NALU even more.
We therefore suggest to leverage the power of the complex plane, by using the
complex logarithm and by promoting $w$ to a complex number.  This will enable
us not only to lift the positivity constraint on $x$, but also provides and
additional dimension that helps to avoid regions with an uninformative gradient
signal\todo{rephrase this. do we have an educating example where it is obvious
that complex plane helps? the stuff in fig.1 does not work as well with
non-complex NPU, but I have not idea how to plot that}.  Introducing a complex
weight matrix somewhere in a larger network would mean that the gradients in
other layers would potentially become complex as well, which would result in
converting the whole network to complex numbers.  This seems like a high price
to pay, but fortunately there is a way around this.  We can output only the
real part of the result $\real(z)$, but still maintain a complex $w$ by
treating it as a sum of real and imaginary part $w=\wre+i\wim$. We are thus
interested in
\begin{align}
  \label{eq:exp_w_log_complex}
  \real(z) &= \real(\exp(w\log x)),
\end{align}
with a real input $x$. The complex logarithm is defined as $\log z=\log r +
i\theta$, which, for real inputs $x$, can be simplified to $\log z = \log r +
ik\pi$, where $k\in \{0,1\}$. With that Eq.~\ref{eq:exp_w_log_complex} can be
rewritten to
\begin{align}
  \real(z) &= \real(\exp((\wre + i\wim) (\log r + ik\pi))) \\
    &= \exp(\wre\log r - \wim k\pi) \cos(\wim\log r + k\pi\wre),
\end{align}
where we used Euler's formula\footnote{Euler's formula states that for any real
number $x$: $re^{ix} = \cos x + i\sin x$}.
\todo{explicitly compute gradients?}

\subsection{Neural Power Unit (NPU)}%
\label{sub:neural_power_unit_npu_}

\begin{figure}
  \centering
  \resizebox{.3\textwidth}{!}{\input{../plots/exp_log_sin/log_comp.tikz.tex}}
  \resizebox{.3\textwidth}{!}{\input{../plots/exp_log_sin/exp_comp.tikz.tex}}
  \resizebox{.3\textwidth}{!}{\input{../plots/exp_log_sin/sin_comp.tikz.tex}}
  \caption{Function learning comparison of different layers.}%
  \label{fig:sin_comp}
\end{figure}

The \emph{Neural Power Unit} (NPU) with two weight matrices $\Wre$ and $\Wim$
then reads
\begin{align}
  \bm r = |\bm x|,
  \hspace{1cm}
  & k_i = \begin{cases}
     0  & x_i \leq 0 \\
    \pi & x_i > 0
  \end{cases}
\end{align}
\begin{align}
  \bm z &= \exp(\Wre \log\bm r - \Wim\bm k) \odot \cos(\Wim\log \bm r + \Wre\bm k)
\end{align}
where $\bm k$ is a vector that is zero where $\bm x$ is positive and $\pi$
where it is negative, $\bm r$ is the element-wise absolute value.
In the final network we are only interested in real powers, so we recommend to
strongly regularize the imaginary part of the weights $\Wim$.

\subsection{Gated NPU}%
\label{sub:gated_npu}
As pointed out by \cite{madsen_neural_2020}, the loss surface of the NALU, and
thus also the NPU, is not always smooth. For some input and weight combinations
it can happen that the loss close to zero becomes very large as shown in
Fig.~\ref{fig:task_loss_npu_nmu}.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{task_loss_npu_nmu.pdf}
  \caption{asdf}%
  \label{fig:task_loss_npu_nmu}
\end{figure}
However, most applications are dealing with batches of data which average out
these cases, as shown in Fig.~\ref{fig:task_loss_npu_nmu_average} for a batch
size of 32.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{task_loss_npu_nmu_average.pdf}
  \caption{asdfadsf}%
  \label{fig:task_loss_npu_nmu_average}
\end{figure}

In practice we encountered difficulties of obtaining sparse solutions with the
NPU. If the inputs to an NPU unit are small, the gradient becomes practically
zero, as we can see in Fig.~\ref{fig:id_loss_npu_nmu}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{id_loss_npu_nmu.pdf}
  \caption{learning identity with small inputs}%
  \label{fig:id_loss_npu_nmu}
\end{figure}

@@

\appendix
\section*{Appendix A}%
\label{sec:appendix_a}

\todo{add sqrt to table~\ref{tab:compare_npu_npu_nalu}?}
\begin{table}[h]
  \centering
  \caption{Comparison of different $2\times2$ layers on the task
  $(x,y)\rightarrow(xy,\frac{x}{y})$.  $(x,y) \in \mathcal U^2(-2,2)$.}
  \label{tab:compare_npu_npu_nalu}
  \input{table-xy-xdivy-pos.tex}
\end{table}
